---
title: "2018 Bacterial Abundance"
author: "Simran Bawa"
date: "10/29/2020"
output: github_document
---

##Intro

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(readxl)
library(lubridate)
```

##Importing Data

```{r}
excel_sheets("~/Desktop/EEMB 144L/144l_students/Input_Data/week3/144L_2018_BactAbund.xlsx")

metadata <- read_excel("~/Desktop/EEMB 144L/144l_students/Input_Data/week3/144L_2018_BactAbund.xlsx", sheet = "Metadata")
glimpse(metadata)
#unique(metadata$Experiment) only one experiment
#unique(metadata$Location) only one location (Campus Point)
#unique(metadata$Bottle) 8 different bottles (A-H)
#unique(metadata$Treatment) four different experiments (control, ash leachate, mud leachate, and glucose nitrate phosphate)

data <- read_excel("~/Desktop/EEMB 144L/144l_students/Input_Data/week3/144L_2018_BactAbund.xlsx", sheet = "Data")
glimpse(data)

joined <- left_join(metadata, data)
names(joined)
#summary(joined)
glimpse(joined)
```

##Prepare Data

```{r}
cells <- joined %>% 
  mutate(Datetime = ymd_hm(Datetime),
         cells = Cells_ml *1000) %>% 
  group_by(Experiment, Treatment, Bottle) %>% 
  mutate(interv = interval(first(Datetime), Datetime),
         hours = as.numeric(interv)/3600,
         days = hours/24) %>% 
  ungroup() %>% 
  select(Experiment:DNA_SampleID, hours, days, cells) %>% 
  drop_na (cells)

glimpse(cells)
```

##Plot Growth Curves

```{r}
levels <- c("Control", "Ash Leachate", "Mud Leachate", "Glucose_Nitrate_Phosphate")

cells %>% 
  mutate(dna = ifelse(DNA_Sample == T, "*", NA)) %>% 
  ggplot(aes(x = days, y = cells, group = interaction(Experiment, Treatment, Bottle))) +
  geom_line(aes(color = factor(Treatment)), size = 1) +
  geom_point(aes(fill = Treatment), size = 3, shape = 21) +
  geom_text(aes(label = dna), size = 12) +
  labs(x = "Days", y = expression(paste("Cells, L"^-1)), fill = "") +
  guides(color = F)
```

## Identify Exponential Phase of Growth

```{r}
ln_cells <- cells %>% 
  group_by(Experiment, Treatment, Bottle) %>% 
  mutate(ln_cells = log(cells), 
         diff_ln_cells = ln_cells - lag(ln_cells, default = first(ln_cells)))
```

```{r}
ln_cells %>% 
  mutate(dna = ifelse(DNA_Sample == T, "*", NA)) %>% 
  ggplot(aes(x = days, y = diff_ln_cells, group = interaction(Experiment, Treatment, Bottle))) +
  geom_line(aes(color = factor(Treatment)), size = 1) +
  geom_point(aes(fill = Treatment), size = 3, shape = 21) +
  geom_text(aes(label = dna), size = 12) +
  labs(x = "Days", y = expression(paste("∆ Ln of Cells, L"^-1)), fill = "") +
  guides(color = F) +
  facet_grid(~Bottle, scales = "free")
```

- SB Bottle A ~0-1 d (T0-T1)
- SB Bottle B ~0-1.25 d (T0-T1)
- SB Bottle C ~0-1 d (T0-T1)
- SB Bottle D ~0-1 d (T0-T1)
- SB Bottle E ~0-1 d (T0-T1)
- SB Bottle F ~0-1 d (T0-T1)
- SB Bottle G ~0-2 d (T0-T2)
- SB Bottle H ~0-2 d (T0-T2)


## Calulcate Growth Rates, Doubling Times, and ∆Cell Abundance

```{r}
growth <- ln_cells %>% 
  mutate(exp_start = 0,
         exp_end = ifelse(Bottle %in% c("G", "H"), 2, 1 )) %>% 
  group_by(Experiment, Treatment, Bottle) %>% 
  mutate(ln_cells_exp_start = ifelse(Timepoint == exp_start, ln_cells, NA), 
         ln_cells_exp_end = ifelse(Timepoint == exp_end, ln_cells, NA), 
         cells_exp_start = ifelse(Timepoint == exp_start, cells, NA), 
         cells_exp_end = ifelse(Timepoint == exp_end, cells, NA), 
         days_exp_start = ifelse(Timepoint == exp_start, days, NA), 
         days_exp_end = ifelse(Timepoint == exp_end, days, NA), ) %>% 
  fill(ln_cells_exp_start:days_exp_end, .direction = "updown") %>% 
  mutate(mew = (ln_cells_exp_end - ln_cells_exp_start)/(days_exp_end - days_exp_start), 
         doubling = log(2)/mew,
         delta_cells = cells_exp_end - first(cells)) %>% 
  ungroup()
glimpse(growth)
```

## Convert Bacterial Abundance & Change in Bacterial Abundance to Carbon Units

```{r}
bactcarbon <- growth %>% 
  mutate(bc = cells * (2.5 * 10^-9), 
         delta_bc = delta_cells * (2.5 * 10^-9))
glimpse(bactcarbon)
```

## Calculate Treatment Averages 

```{r}
averages <- bactcarbon %>%
  group_by(Experiment, Treatment, Timepoint) %>% 
  mutate(ave_bc = mean(bc)) %>% 
  ungroup() %>% 
  group_by(Experiment, Treatment) %>% 
  mutate(ave_mew = mean(mew),
         ave_doubling = mean(doubling),
         ave_delta_cells = mean(delta_cells),
         ave_delta_bc = mean(delta_bc),
         ave_lag = mean(days_exp_start),
  ) %>% 
  ungroup()
  
glimpse(averages)  
```

## Plot Treatment Averages

```{r fig.height=4, fig.width=6}
averages %>% 
  ggplot(aes(x = days, y = ave_bc), group = interaction(Experiment, Treatment)) +
  geom_errorbar(aes(ymin = ave_bc, ymax = ave_bc, color = factor(Treatment, levels = levels)), width = 0.1) +
  geom_line(aes(color = factor(Treatment, levels = levels)), size = 1) +
  geom_point(aes(fill = factor(Treatment, levels = levels)), color = "black", shape = 21, size = 3) +
  facet_grid(rows = "Location", scales = "free") +
  labs(x = "Days", y = expression("Bacterial Carbon, µmol C L"^-1), fill = "", color = "") +
  guides(color = F) +
  theme_bw()
```

## Barplots

```{r}
bar.data <- averages %>% 
  select(Location, Treatment, ave_mew:ave_lag) %>% 
  distinct()
```

```{r}
mew <- bar.data %>% 
  ggplot(aes(x = factor(Treatment, levels = levels), y = ave_mew), group = interaction(Location, Treatment)) +
  geom_col(color = "black", fill = "white") +
  geom_errorbar(aes(ymin = ave_mew, ymax = ave_mew), width = 0.1) +
  facet_grid(~factor(Location, levels = levels), scales = "free") +
  labs(x = "", y = expression("µ, d"^-1)) +
  theme_bw()
```


```{r}
doubling <- bar.data %>% 
  ggplot(aes(x = factor(Treatment, levels = levels), y = ave_doubling), group = interaction(Location, Treatment)) +
  geom_col(color = "black", fill = "white") +
  geom_errorbar(aes(ymin = ave_doubling, ymax = ave_doubling), width = 0.1) +
  facet_grid(~factor(Location, levels = levels), scales = "free") +
  labs(x = "", y = expression("Doubling Time, d")) +
  theme_bw()
```


```{r}
delta_bc <- bar.data %>% 
  ggplot(aes(x = factor(Treatment, levels = levels), y = ave_delta_bc), group = interaction(Location, Treatment)) +
  geom_col(color = "black", fill = "white") +
  geom_errorbar(aes(ymin = ave_delta_bc, ymax = ave_delta_bc), width = 0.1) +
  facet_grid(~factor(Location, levels = levels), scales = "free") +
  labs(x = "", y = expression("∆ Bacterial Carbon, µmol C L"^-1)) +
  theme_bw()
```

```{r}
lag <- bar.data %>% 
  ggplot(aes(x = factor(Treatment, levels = levels), y = ave_lag), group = interaction(Location, Treatment)) +
  geom_col(color = "black", fill = "white") +
  geom_errorbar(aes(ymin = ave_lag, ymax = ave_lag), width = 0.1) +
  facet_grid(~factor(Location, levels = levels), scales = "free") +
  labs(x = "", y = "Lag Phase, days") +
  theme_bw()
```

```{r}
#install.packages("patchwork")
library(patchwork)
```

```{r fig.height=6, fig.width=8}
lag + delta_bc + mew + doubling + plot_annotation(tag_levels = "a")
```


# Save Data 

```{r}
saveRDS(averages, "~/Desktop/EEMB 144L/144l_students/BactAbund.rds")
```















